Why there are more species in several small patches vs few large patches:

### 2024-03-08 ms_SLOSS.Rmd
Correciones menores

Leonardo A. Saravia ^1^, Ulises Balza ^1^, Fernando Momo^2^

1. Centro Austral de Investigaciones Científicas (CADIC-CONICET). Houssay 200. Ushuaia, Argentina

2. Instituto de Ciencias. Universidad Nacional de General Sarmiento. J.M. Gutierrez 1159 (1613), Los Polvorines.        Buenos Aires, Argentina.

@@

The SLOSS debate has been approached from several theoretical frameworks, but most of them were based on single-species models [@Fahrig2021]. <!--Moreover, spatial dimension in several works has been neglected as mere noise around long-term equilibrium dynamics [cita?]-->. However, recent studies using spatially explicit simulations revealed that while high fragmentation leads to more local extinctions due to the reduced habitat size, at low levels of fragmentation, competitive exclusion is more prevalent, resulting in decreased diversity [@Rybicki2020].

By examining spatially explicit models with a fixed level of habitat destruction alongside neutral and competitive community dynamics, encompassing diverse dispersal-related traits and various geometries of habitat loss, we aim to align the model parameters with the recently proposed SLOSS cube hypothesis  [@Fahrig2021]. This hypothesis serves as a comprehensive framework summarizing existing SLOSS predictions and empirical findings. According to the SLOSS cube hypothesis, the condition where SL > SS will prevail is contingent upon the simultaneous occurrence of the following conditions: low between-patch movement rate, minimal impact of spreading-of-risk on population dynamics, and low species spatial clumping [@Fahrig2021].

This work aims to explore spatial multi-species models to verify the conditions when SS > SL pattern holds and determine the causes of it. In particular, we simulate species retention after different fragmentation scenarios of the same overall habitat loss, therefore accounting for the interacting effects of habitat loss and fragmentation.

@@

@@

Table: Parameter values used in simulations. Factorial combinations of these parameters were used to run simulations. $\lambda$ represents the ratio of growth rate to death rate. "Dispersal distance" refers to the mean of the power law dispersal kernel. "Habitat patch size" indicates the side length of habitat patches modeled as squares.

@@

Naturally, there are limitations to our modeling work. Firstly, it is important to note that more research is needed to apply our to determine the effect of fragmentation in real-world communities. We assume that all species have the same parameters, thus hypothetically determining these parameters for all the species, and using the mean to determine the responses to fragmentation would render an approximate value for the effects and thresholds. Also, an improved version of the model with different values of the parameters for each species could be developed and compared with the neutral model. The effects of fragmentation could be influenced by many other ecological/behavioural characteristics of the species not considered in the model, including environmental heterogeneity, matrix permeability, and individual’s cost of dispersal, among others. Further, the size of the focal landscape compared to dispersal distances, which determines the scale of the dynamics, could influence quantitative results [@Rybicki2020]. Besides that, the approach of using mean values of the parameters of relatively simple multispecies models has been applied and tested to predict species biomass [@Fort2018].

>>>>

Leonardo A. Saravia ^1^^2^, Ulises Balza ^1^, Fernando Momo^3^^4^

1. Centro Austral de Investigaciones Científicas (CADIC-CONICET). Houssay 200 (9410), Ushuaia, Argentina.

2. Universidad Nacional de Tierra del Fuego (UNTDF), Yrigoyen 879 (9410), Ushuaia, Argentina

3. Instituto de Ciencias. Universidad Nacional de General Sarmiento. J.M. Gutierrez 1159 (1613), Los Polvorines, Argentina.

4. Departamento de Ciencias Básicas. Universidad Nacional de Luján. Rutas 5 y 7, (6700) Luján. Argentina.

@@

The SLOSS debate has been approached from several theoretical frameworks, but most of them were based on single-species models [@Fahrig2021]. However, recent studies using spatially explicit simulations revealed that while high fragmentation leads to more local extinctions due to the reduced habitat size, at low levels of fragmentation, competitive exclusion is more prevalent, resulting in decreased diversity [@Rybicki2020].

The recently proposed SLOSS cube hypothesis by @Fahrig2021 serves as a comprehensive framework summarizing existing Single Large or Several Small (SLOSS) predictions and empirical findings. This hypothesis posits that the condition where Single Large (SL) patches contribute more to biodiversity conservation than Several Small (SS) patches is contingent upon the simultaneous occurrence of specific conditions: a low between-patch movement rate, minimal impact of spreading-of-risk on population dynamics, and low species spatial clumping [@Fahrig2021].

This research aims to investigate multi-species spatially explicit models, incorporating a fixed level of habitat destruction, neutral and competitive community dynamics, diverse dispersal-related traits, and various geometries of habitat loss. The principal objective is to verify the conditions under which the SS > SL pattern holds and discern the underlying causes. Through this examination, we seek to contribute insights that can help determine if the conditions outlined in the SLOSS cube hypothesis are supported by our modeling approach..

Specifically, our approach involves simulating species retention under diverse fragmentation scenarios, all characterized by a fixed amount of habitat loss. We vary parameters such as patch sizes, community type (competitive or neutral), species dispersal capabilities, and the community's persistence status—whether it is near extinction and degraded or in a healthy state. This comprehensive exploration enables us to account for the interacting effects of habitat loss and a range of model parameters on species retention.

@@

Table: Parameter values utilized in simulations, with factorial combinations of these parameters employed to run simulations. The symbol $\lambda$ represents the ratio of growth rate to death rate. "Dispersal distance" denotes the mean of the power law dispersal kernel, while "Habitat patch size" indicates the side length of habitat patches modeled as squares. Both the dispersal distance and habitat patch size are measured in grid cells.

@@

@@

Naturally, there are limitations to our modeling work. Firstly, it is important to note that more research is needed to apply our to determine the effect of fragmentation in real-world communities. We assume that all species have the same parameters, thus hypothetically determining these parameters for all the species, and using the mean to determine the responses to fragmentation would render an approximate value for the effects and thresholds. Also, an improved version of the model with different values of the parameters for each species could be developed and compared with the neutral model. The effects of fragmentation could be influenced by many other ecological/behavioural characteristics of the species not considered in the model, including environmental heterogeneity, matrix permeability, and individual’s cost of dispersal, among others. Further, the size of the focal landscape compared to dispersal distances, which determines the scale of the dynamics, could influence quantitative results [@Rybicki2020]. Additionally, the application of mean parameter values in relatively simple multispecies models has been successfully employed and tested for predicting species biomass, as demonstrated by @Fort2018.
